Probabilistic Model of Bilateral Lymphatic Spread in Head and Neck Cancer

### 2024-08-23 _tex/manuscript.tex
Built site for gh-pages

\raggedbottom
\usepackage{multirow}
\usepackage{centernot}
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\usepackage{bookmark}

@@

\subsection{Consensus on Involvement
Status}\label{consensus-on-involvement-status}

@@

These datasets allow us to investigate correlations between the
involvement of individual LNLs, or between risk factors and patterns of
involvement. In figure~\ref{fig-data-strat}, we have plotted the
prevalence of each contralateral LNL's involvement, stratified by
T-category, ipsilateral number of involved LNLs, and whether the tumor
extended over the mid-sagittal line. A similar but more complete
stratification is also tabulated in the appendix in
table~\ref{tbl-data-strat}.

@@

The three risk factors for contralateral involvement, midline extension,
T-category, and ipsilateral involvement are correlated. For example,
midline extension was present in 48\% of advanced T-category tumors but
only in 7\% of early T-category tumors. The higher prevalence of
contralateral metastases in advanced T-category tumors may thus be
partially explained by a higher rate of midline extension in these
tumors. However, figure~\ref{fig-data-strat-uncorr} indicates that
advanced T-category and severe ipsilateral involvement do represent
additional risk factors . The figure considers only patients with
lateralized tumors not extending over the midline. For early T-category
and healthy ipsilateral levels I-V, only 1\% (1 out of 84 patients)
shows involvement of contralateral level II. This increases to 5\% (9
out of 165 patients) if ipsilateral level II is involved, to 15\% (9 out
of 60 patients) if ipsilateral levels II and III are involved, and to
23\% (6 out of 26 patients) for advanced T-category with ipsilateral
levels II and III involved.

@@

Based on the observations of the \hyperref[sec-data-strat]{previous
section}, any potential model that aims to also predict the risk for
contralateral nodal involvement, should be able to take the following
into account:

@@

Over the course of this work, we will first briefly recap the mentioned
HMM in section~\ref{sec-unilateral}, which was so far used to model
ipsilateral lymphatic progression only. Then, we intuitively extend it
to include the contralateral side as well in
section~\ref{sec-ext-to-contra}. In this section, we also introduce a
way of modelling the tumor's midline extension as a random variable
(section~\ref{sec-midline}) and lastly talk about how it may affect the
contralateral spread in section~\ref{sec-params-symmetry}.

@@

\subsection{Likelihood Function}\label{likelihood-function}

@@

In general, the matrices \(\boldsymbol{\Lambda}_\text{i}\) and
\(\boldsymbol{\Lambda}_\text{c}\) could be parameterized using a
disjoint set of parameters. I.e., the ipsi- and contralateral spread
rates are entirely different. However, using three sensible assumptions,
we can reduce the parameter space by sharing some parameters between the
sides:

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  The tumor was still lateralized at \(t=\tau\) and just grew over the
  midline. Under this scenario, the probability for having a midline
  extension at \(t=\tau+1\) is given by \(p_\epsilon\). We use this to
  weight the contralateral state distribution that was so far evolved
  without increased contralateral spread.
\item
  The mid-sagittal line was already crossed by the tumor. In that case,
  the probability is 1 to remain in that lateralization state. Thus, for
  this case we simply add the same distribution we are computing, but
  from one time-step earlier, arriving at a recursion relation:
\end{enumerate}

@@

It is possible to provide our model with multiple different diagnostic
modalities, each being characterized by different pairs of sensitivity
and specificity. However, we instead chose to combine them into a single
``consensus'' diagnosis before parameter inference. We opted for this
because the literature values of sensitivity and specificity
\citep{de_bondt_detection_2007, kyzas_18f-fluorodeoxyglucose_2008} of
imaging modalities like MRI and CT do not plausibly match some of our
observations: In the USZ cohort, 78\% of OPSCC patients where diagnosed
with ipsilateral LNL II involvement via diagnostic imaging. This is
virtually impossible with sensitivities around 80\% and specificities
lower than 100\%.

Our pre-training consensus was formed by considering all reported
diagnostic information for a particular patient and LNL. When conflicts
arose, we computed the \emph{most likely} true state of involvement
using the literature sensitivity and specificity values
\citep{de_bondt_detection_2007, kyzas_18f-fluorodeoxyglucose_2008}.

>>>>

\raggedbottom
\usepackage{multirow}
\usepackage{centernot}
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother

\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\usepackage{bookmark}

@@

\subsection{Consensus on Involvement Status}\label{sec-data-consensus}

@@

These datasets allow us to investigate correlations between the
involvement of individual LNLs, or between risk factors and involvement.
In figure~\ref{fig-data-strat}, we have plotted the prevalence of each
contralateral LNL's involvement, stratified by T-category, number of
ipsilaterally involved LNLs, and whether the tumor extended over the
mid-sagittal line. A similar but more complete stratification is also
tabulated in the appendix in table~\ref{tbl-data-strat}.

@@

The three risk factors for contralateral involvement, midline extension,
T-category, and ipsilateral involvement are correlated. For example,
midline extension was present in 48\% of advanced T-category tumors but
only in 7\% of early T-category tumors. The higher prevalence of
contralateral metastases in advanced T-category tumors may thus be
partially explained by a higher rate of midline extension in these
tumors. However, figure~\ref{fig-data-strat-uncorr} indicates that
advanced T-category and severe ipsilateral involvement do represent
additional risk factors. The figure considers only patients with
lateralized tumors not extending over the midline. For early T-category
and healthy ipsilateral levels I-V, only 1\% (1 out of 84 patients)
shows involvement of contralateral level II. This increases to 5\% (9
out of 165 patients) if ipsilateral level II is involved, to 15\% (9 out
of 60 patients) if ipsilateral levels II and III are involved, and to
23\% (6 out of 26 patients) for advanced T-category with ipsilateral
levels II and III involved.

@@

Based on the observations of the \hyperref[sec-data-strat]{previous
section}, any model to predict the risk for contralateral nodal
involvement, should be able to describe the following:

@@

Over the course of this work, we will first briefly recap the HMM in
section~\ref{sec-unilateral}, which was so far used to model ipsilateral
lymphatic progression only. Then, we intuitively extend it to include
the contralateral side as well in section~\ref{sec-ext-to-contra}. In
this section, we also introduce a way of modelling the tumor's midline
extension as a random variable (section~\ref{sec-midline}) and lastly
talk about how it may affect the contralateral spread in
section~\ref{sec-params-symmetry}.

@@

\subsection{Likelihood Function of the Unilateral
Model}\label{likelihood-function-of-the-unilateral-model}

@@

In general, the matrices \(\boldsymbol{\Lambda}_\text{i}\) and
\(\boldsymbol{\Lambda}_\text{c}\) could be parameterized using a
disjoint set of parameters, i.e., the ipsi- and contralateral spread
rates are entirely different. However, using three sensible assumptions,
we can reduce the parameter space by sharing some parameters between the
two sides:

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  The tumor was still lateralized at \(t=\tau\) and just grew over the
  midline. Under this scenario, the probability for having a midline
  extension at \(t=\tau+1\) is given by \(p_\epsilon\). We use this to
  weight the contralateral state distribution that was so far evolved
  without increased contralateral spread.
\item
  The mid-sagittal line was already crossed by the tumor. In that case,
  the probability is 1 to remain in that lateralization state. Thus, to
  account for this case we simply add the distribution
  \(P\left( \mathbf{X}^\text{c}, \epsilon=\texttt{True} \mid \tau \right)\)
  from one time-step earlier, arriving at a recursion relation:
\end{enumerate}

@@

It is possible to provide our model with multiple different diagnostic
modalities, each being characterized by different pairs of sensitivity
and specificity. However, we instead chose to combine them into a single
``consensus'' diagnosis before parameter inference, as introduced in
\textbf{?@sec-sec-data-consensus}. We opted for this because the
literature values of sensitivity and specificity
\citep{de_bondt_detection_2007, kyzas_18f-fluorodeoxyglucose_2008} of
imaging modalities like MRI and CT do not plausibly match some of our
observations: In the USZ cohort, 78\% of OPSCC patients where diagnosed
with ipsilateral LNL II involvement via diagnostic imaging. This is
virtually impossible with sensitivities around 80\% and specificities
lower than 100\%.
